Abstractions for Multi-Sorted Substitutions

### 2024-03-14 paper/paper.tex
Update.

  % In summary, to have enough substitution machinery to prove subject
  % reduction for a language with expression-, type-, and
  % kind-variables one can end up with a total of 24
  % repetitions for each definition related to substitution, not counting
  % the interaction lemmas between different substitutions that are
  % acting on the same syntactic categories!
  If one is not careful, a formalization of a language with type and
  kind polymorphism can easily end up with 24 slightly changed copies of
  the whole substitution machinery, which for de Bruijn indices tends to
  include at least a type for parallel substitution, and operations for
  application, extension, weakening, lifting, singleton substitution,
  extensional equality, and an identity substitution - and this is not
  counting the slightly changed copies of interaction lemmas!
  Clearly, this situation is in need of automation.

  Our framework approaches this problem using a combination of
  abstractions and reflection in the context of extrinsically typed,
  intrinsically scoped syntax with de Bruijn indices.
  The user can write a language specification using regular Agda
  definitions (no generic encodings) and our framework derives
  definitions and lemmas for untyped substitution, and provides an
  abstraction to prove type preservation for all substitution operations
  using a single lemma. For System F, our framework allows to prove
  subject reduction with only a single handwritten lemma for type
  preservation.

@@

  The rest of this paper will introduce our framework using System F
  as a running example for a type soundness proof.

@@

  In this paper we present a simplified version of the actual
  framework, focusing on the core concepts. We present all necessary
  definitions and lemmas, but omit some proofs for the sake of
  space. The omitted proofs can be found in Appendix~\ref{sec:proofs}
  and in the supplementary material.
  The full framework is available online.\footnote{See supplementary material. In the final version, this text will be replaced by a link to github.}

  The rest of this paper is structured as followed:
  Section~\ref{sec:syntax} introduces the multi-sorted syntax and compares it to the
  more common unsorted syntax.
  Section~\ref{sec:maps} introduces multi-sorted substitutions and renamings, and
  an abstraction to unify them.
  Section~\ref{sec:composition} introduces composition of multi-sorted
  substitions and renamings, and an abstraction to unify all four
  compositions.
  Section~\ref{sec:types} shows how to introduce the notion of types
  for multi-sorted syntax and how to define typing contexts.
  Section~\ref{sec:typing} shows how to define a multi-sorted typing relation,
  and an abstraction to unify type preservation lemmas for renamings and substitutions.
  % Section~\ref{sec:subjectreduction} introduces a small-step
  % operational semantics for System F and proves subject reduction.
  Section~\ref{sec:reflection} describes the class of object languages covered by
  our reflection algorithm.
  Section~\ref{sec:casestudies} discusses our case studies.
  Section~\ref{sec:related} discusses related work.
  Section~\ref{sec:conclusion} concludes with future work.

  \subsection{Contributions}
  \label{sec:introduction:contributions}
  \begin{enumerate}
  \item
    a novel approach for formalizing intrinsically-scoped syntax with
    multiple variable sorts as a special kind of intrinsically-typed
    syntax, we call \emph{multi-sorted syntax};
  \item
    an adaptation of McBride's
    kit-abstraction\cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}
    for unifying renamings and substitutions to multi-sorted syntax;
  \item
    a novel abstraction for composition and it's metatheory, unifying
    the four compositions between renamings and substitutions;
  \item
    a novel abstraction for typing relations, unifying type
    preservation of renaming and substitution;
  \item
    a formalized specification of a large class of object languages
    for which untyped substitution and lemmas can be derived generically.
  \item
    an implementation as an Agda framework featuring
    a reflection algorithm, representation independence for
    substitutions and typing contexts, heterogeneous equality between
    renamings and substitutions, and absence of postulated axioms.
  \item
    three case studies in using our approach to prove subject reduction
    for System~F with subtyping, a dependently-typed lambda calculus, and
    pattern matching.
  \end{enumerate}

>>>>

  % In summary, to have enough substitution machinery to prove subject
  % reduction for a language with expression-, type-, and
  % kind-variables one can end up with a total of 24
  % repetitions for each definition related to substitution, not counting
  % the interaction lemmas between different substitutions that are
  % acting on the same syntactic categories!
  If one is not careful, a formalization of a language with type and
  kind polymorphism can easily end up with 24 slightly changed copies of
  the whole substitution machinery!
  Clearly, this situation is in need of automation.

  Our framework approaches this problem by using a combination of
  abstractions and reflection in the context of extrinsically typed,
  intrinsically scoped syntax with de Bruijn indices.
  The user can write a language specification using regular Agda
  definitions (no generic encodings) and our framework derives
  definitions and lemmas for untyped substitution, and provides an
  abstraction to prove type preservation for all substitution operations
  with only a single lemma. For System F, our framework allows to prove
  subject reduction with only a single handwritten lemma for type
  preservation.

@@

  The rest of this paper will introduce our framework using System F
  as a running example for a substitution-preserves-typing proof.

@@

  In this paper we present a simplified version of the actual
  framework, focusing on the core concepts. We present all necessary
  definitions and lemmas, but omit some proofs for the sake of
  space. The omitted proofs can be found in Appendix~\ref{sec:proofs}
  and in the supplementary material.
  The full framework is available on Github: \textcolor{blue}{\url{https://github.com/m0rphism/kitty}}
  % The full framework is available online.\footnote{See supplementary material. In the final version, this text will be replaced by a link to github.}

  The rest of this paper is structured as follows:
  Section~\ref{sec:syntax} introduces the multi-sorted syntax and compares it to the
  more common unsorted syntax.
  Section~\ref{sec:maps} introduces multi-sorted substitutions and renamings, and
  an abstraction to unify them.
  Section~\ref{sec:composition} introduces composition of multi-sorted
  substitions and renamings, and an abstraction to unify all four
  compositions.
  Section~\ref{sec:types} shows how to define types
  and typing contexts for multi-sorted syntax.
  Section~\ref{sec:typing} shows how to define a multi-sorted typing relation,
  and an abstraction to unify type preservation lemmas for renamings and substitutions.
  % Section~\ref{sec:subjectreduction} introduces a small-step
  % operational semantics for System F and proves subject reduction.
  Section~\ref{sec:reflection} describes the class of object languages covered by
  our reflection algorithm.
  Section~\ref{sec:casestudies} discusses our case studies.
  Section~\ref{sec:related} discusses related work.
  Section~\ref{sec:conclusion} concludes.

  \subsection{Contributions}
  \label{sec:introduction:contributions}
  \begin{enumerate}
  \item
    a novel approach for formalizing intrinsically-scoped syntax with
    multiple variable sorts as a special kind of intrinsically-typed
    syntax, we call \emph{multi-sorted syntax};
  % \item
  %   an adaptation of McBride's
  %   kit-abstraction\cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}
  %   for unifying renamings and substitutions to multi-sorted syntax;
  \item
    a novel abstraction for composition and it's metatheory, unifying
    the four compositions between renamings and substitutions;
  \item
    a novel abstraction for typing, unifying type
    preservation of renaming and substitution;
  \item
    a formalized specification of a large class of object languages
    for which untyped substitution and lemmas can be derived generically.
  \item
    an implementation as an Agda framework featuring
    a reflection algorithm, representation independence for
    substitutions and typing contexts, heterogeneous equality between
    renamings and substitutions, and absence of postulated axioms.
  \item
    three case studies in using our approach to prove subject reduction
    for System~F with subtyping, a dependently-typed lambda calculus, and
    pattern matching.
  \end{enumerate}
